---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
pkgload::load_all(quiet = TRUE)
```

```{r quality, include = FALSE}
# Dynamic quality metrics — recomputed on every render so the badges,
# the "Available tools (N)" header and the per-file test counts in the
# repository tree never drift from reality.
n_tools <- length(entsoeapi.mcp:::all_tools())

# Count tests per file without emitting reporter output.
.count_tests <- function(file) {
  res <- testthat::test_file(
    path     = file.path("tests/testthat", file),
    reporter = "silent"
  )
  length(res)
}
n_tests_run   <- .count_tests("test-run.R")
n_tests_tools <- .count_tests("test-tools.R")
n_tests_utils <- .count_tests("test-utils.R")
n_tests_total <- n_tests_run + n_tests_tools + n_tests_utils

# Coverage — round to nearest integer for the badge.
cov_pct <- round(covr::percent_coverage(covr::package_coverage(path = ".")))
```

# entsoeapi.mcp

<!-- badges: start -->

![Tests](https://img.shields.io/badge/tests-%60r%20n_tests_total%60%20passing-brightgreen) ![Coverage](https://img.shields.io/badge/coverage-%60r%20cov_pct%60%25-brightgreen) ![Tools](https://img.shields.io/badge/tools-%60r%20n_tools%60-blue) ![R](https://img.shields.io/badge/R-%E2%89%A5%204.2.0-blue) ![License](https://img.shields.io/badge/license-MIT-blue)

<!-- badges: end -->

An [MCP](https://modelcontextprotocol.io) server that exposes the [ENTSO-E Transparency Platform](https://transparency.entsoe.eu) as tools for LLM-enabled applications such as Claude Desktop. Query European electricity market, load, generation, transmission, outages, and balancing data through natural language.

Built in pure R on top of [`entsoeapi`](https://github.com/krose/entsoeapi) and [`mcptools`](https://github.com/posit-dev/mcptools). Distributed as a Docker image — no R installation required for end users.

------------------------------------------------------------------------

## What you can ask

> *"What was the actual electricity load in Germany on 3 January 2025?"*\
> *"Show me day-ahead prices for France and Germany last Monday."*\
> *"Which generation units in Belgium had unplanned outages this week?"*\
> *"What is the installed solar capacity in the Netherlands in 2024?"*\
> *"Get the net transfer capacity from Germany to Poland for today."*

------------------------------------------------------------------------

## Prerequisites

| Requirement | Purpose | Notes |
|------------------------|------------------------|------------------------|
| [Docker](https://docs.docker.com/get-docker/) | Runs the server container | Any version that supports `docker run`. |
| [Claude Desktop](https://claude.ai/download) | MCP client | or any MCP-compatible client |
| [ENTSO-E API token](https://transparency.entsoe.eu/usrm/user/createPublicUser) | Access to the Transparency Platform | Free registration; UUID v4 format. |

------------------------------------------------------------------------

## Quick start

### Step 1 — Get an ENTSO-E API token

1.  Register at [transparency.entsoe.eu](https://transparency.entsoe.eu/usrm/user/createPublicUser)
2.  After logging in, go to **My Account Settings → Web API Security Token**
3.  Generate a token — it looks like [`xxxxxxxx-xxxx-xxxx-xxxx-xxxxxxxxxxxx`](https://en.wikipedia.org/wiki/Universally_unique_identifier)

### Step 2 — Pull the Docker image using CLI

``` bash
docker pull sbudai/entsoeapi.mcp:latest
```

Or build from source (see [Building from source](#building-from-source) below).

### Step 3 — Configure MCP server settings of Claude Desktop

Open (or create) your Claude Desktop config file:

-   **macOS / Linux**: `~/.config/Claude/claude_desktop_config.json`
-   **Windows**: `%APPDATA%\Claude\claude_desktop_config.json`

Add the `entsoe` entry under `"mcpServers"`:

``` json
{
  "mcpServers": {
    "entsoe": {
      "command": "docker",
      "args": [
        "run", "--rm", "-i",
        "-e", "ENTSOE_PAT",
        "sbudai/entsoeapi.mcp:latest"
      ],
      "env": {
        "ENTSOE_PAT": "your-entsoe-api-token-here"
      }
    }
  }
}
```

Key flags:\
- `--rm` — removes the container when the session ends (no container accumulation)\
- `-i` — keeps stdin open, required for MCP stdio transport\
- `-e ENTSOE_PAT` — forwards the token from Claude Desktop into the container\
- No `-t` — TTY mode would corrupt the JSON-RPC stream

### Step 4 — Restart Claude Desktop and verify

Restart Claude Desktop to load the new MCP server configuration. You should see **entsoe** listed under the tools panel.

Test with a simple query:

> *"What is the EIC code for Germany?"*

Claude should call the `area_eic` tool and return a list of German bidding zones and control areas.

------------------------------------------------------------------------

## Available tools (`r n_tools`)

All tools return results as CSV (with a truncation notice if capped at 100 rows). All date parameters use **CET timezone** and **YYYY-MM-DD format**.

### EIC lookup

| Tool | Description |
|------------------------------------|------------------------------------|
| `area_eic` | Bidding zones, control areas, member states — look up EIC codes by country name |
| `party_eic` | Market participants (TSOs, traders, etc.) |
| `all_approved_eic` | All approved EIC codes across all types |
| `resource_object_eic` | Generation units and other resource objects |
| `get_news` | Latest news and maintenance notices from ENTSO-E |

### Load

One tool with a `type` parameter:

| `type` value         | ENTSO-E dataset | Description                           |
|--------------------|------------------|-----------------------------------|
| `actual` *(default)* | 6.1.A           | Actual total load                     |
| `day_ahead`          | 6.1.B           | Day-ahead total load forecast         |
| `week_ahead`         | 6.1.C           | Week-ahead total load forecast        |
| `month_ahead`        | 6.1.D           | Month-ahead total load forecast       |
| `year_ahead`         | 6.1.E           | Year-ahead total load forecast        |
| `margin`             | 8.1             | Year-ahead generation forecast margin |

### Generation

One tool (`gen_time_series`) with a `type` parameter:

| `type` value | ENTSO-E dataset | Description |
|------------------------|------------------------|------------------------|
| `actual` *(default)* | 16.1.B&C | Actual generation per production type |
| `wind_solar` | 14.1.D | Day-ahead wind and solar forecast |
| `day_ahead` | 14.1.C | Day-ahead aggregated generation forecast |
| `per_unit` | 16.1.A | Actual generation per generation unit *(slow)* |
| `storage` | 16.1.D | Hydro reservoir / water storage filling rate |

### Generation — installed capacity

One tool (`gen_capacity`) with a `per_unit` parameter:

| `per_unit` value    | ENTSO-E dataset | Description                      |
|---------------------|-----------------|----------------------------------|
| `FALSE` *(default)* | 14.1.A          | aggregate by production type     |
| `TRUE`              | 14.1.C          | individual production unit level |

### Market

| Tool | ENTSO-E dataset | Description |
|------------------------|------------------------|------------------------|
| `energy_prices` | 12.1.D | Day-ahead / intraday market clearing prices |
| `intraday_prices` | 12.1.D | Intraday market prices |
| `net_transfer_capacities` | 11.1 | NTC between two bidding zones |
| `day_ahead_commercial_sched` | 12.1.F | Day-ahead commercial schedules |
| `explicit_offered_transfer_capacities` | 12.1.A | Explicit offered transfer capacities |
| `flow_based_allocations` | 12.1.B | Flow-based allocations for CWE/Core region |
| `congestion_income` | 12.1.G | Congestion income |
| `allocated_transfer_capacities_3rd_countries` | 12.1.C | Allocated transfer capacities on third-country borders |

### Transmission

| Tool | ENTSO-E dataset | Description |
|------------------------|------------------------|------------------------|
| `cross_border_physical_flows` | 12.1.G | Physical cross-border flows |
| `total_commercial_sched` | 12.1.F | Total commercial schedules |
| `net_positions` | 12.1.H | Net positions |
| `forecasted_transfer_capacities` | 11.1 | Forecasted transfer capacities |

### Outages

| Tool | ENTSO-E dataset | Description |
|------------------------|------------------------|------------------------|
| `outages_gen_units` | 15.1.A&B | Unavailability of generation units |
| `outages_prod_units` | 15.1.C&D | Unavailability of production units |
| `outages_transmission_grid` | 15.1.A&B | Unavailability in the transmission grid |

### Balancing

| Tool                  | ENTSO-E dataset | Description                   |
|-----------------------|-----------------|-------------------------------|
| `imbalance_prices`    | 17.1.D          | Imbalance prices              |
| `imbalance_volumes`   | 17.1.C          | Imbalance volumes             |
| `contracted_reserves` | 17.1.B          | Contracted balancing reserves |

------------------------------------------------------------------------

## Tips for good results

-   **Always call `area_eic` first** to resolve a country name to its EIC code before querying data. Many countries have multiple bidding zones (e.g. Norway has NO1–NO5).
-   **Use short date ranges** (1–7 days). ENTSO-E data is at 15-minute to 1-hour resolution — a full week produces \~672 rows. Results are capped at 100 rows with a truncation notice.
-   **For generation time-series**, omit `gen_type` to get all production types, or filter with codes like `B16` (Solar) or `B18` (Wind offshore).

## Data notes

-   **Timezone**: All dates are interpreted as CET (Central European Time; UTC+1 / UTC+2 in summer) regardless of your local timezone.
-   **Resolution**: Load and generation data is typically at 15-minute or 1-hour intervals
-   **Row cap**: Results are capped at 100 rows per call. A truncation notice is appended when data is cut. Use short date ranges (1–7 days) to avoid truncation
-   **Max range**: Most time-series tools accept up to 1 year per query
-   **EIC codes**: Use `area_eic` to resolve country or zone names to EIC codes before querying data

------------------------------------------------------------------------

## Alternative: native R (no Docker)

If you already have R ≥ 4.2.0 installed:

``` bash
# Install dependencies
Rscript -e "install.packages('pak'); pak::pkg_install(c('krose/entsoeapi', 'posit-dev/mcptools', 'ellmer', 'lubridate', 'jsonlite'))"

# Install this package
R CMD INSTALL /path/to/entsoeapi.mcp
```

Then in `claude_desktop_config.json`:

``` json
{
  "mcpServers": {
    "entsoe": {
      "command": "Rscript",
      "args": ["-e", "entsoeapi.mcp::run()"],
      "env": {
        "ENTSOE_PAT": "your-entsoe-api-token-here"
      }
    }
  }
}
```

------------------------------------------------------------------------

## Building from source {#building-from-source}

``` bash
git clone https://github.com/sbudai/entsoeapi.mcp.git
cd entsoeapi.mcp

# Build the Docker image
docker build -t sbudai/entsoeapi.mcp:latest .

# Verify the server starts and responds to MCP initialize
echo '{"jsonrpc":"2.0","id":1,"method":"initialize","params":{"protocolVersion":"2024-11-05","capabilities":{},"clientInfo":{"name":"test","version":"0"}}}' \
  | docker run --rm -i -e ENTSOE_PAT=test sbudai/entsoeapi.mcp:latest
```

The image is based on [`rocker/r-ver:4.4.3`](https://rocker-project.org), uses [Posit Public Package Manager](https://packagemanager.posit.co) for pre-built Linux binaries, and runs as a non-root `mcp` user.

------------------------------------------------------------------------

## Testing

The package ships with **`r n_tests_total` unit tests** that exercise every tool wrapper, every dispatch branch, and every utility — yielding **`r cov_pct`% line coverage** across `R/run.R`, `R/tools.R`, and `R/utils.R`.

Run the suite locally:

``` r
# Plain run (fast)
pkgload::load_all()
testthat::test_dir("tests/testthat")

# With coverage report
covr::package_coverage() |> covr::report()
```

Tests are network-free: every `entsoeapi::*` call is mocked at the package's *imports environment* via a small `.with_imp_bindings()` helper, and `safe_to_csv()` is mocked in the namespace via `testthat::with_mocked_bindings(.package = "entsoeapi.mcp")`. This lets the suite run in CI without an `ENTSOE_PAT` token and without hitting the Transparency Platform.

------------------------------------------------------------------------

## Repository structure

```         
entsoeapi.mcp/
├── Dockerfile            # Self-contained runtime image
├── DESCRIPTION           # Package metadata and dependencies
├── NAMESPACE             # Exported functions (generated by roxygen2)
├── LICENSE               # MIT
├── NEWS.md               # What has changed in each version
├── README.md             # GitHub standard version of README.Rmd
├── README.Rmd            # Comprehensive introduction of the project
├── RELEASE_CHECKLIST     # What to check during launching a new version
├── R/
|   ├── run.R             # Entry point: run() → mcp_server(tools = all_tools())
|   ├── tools.R           # All `r n_tools` ellmer::tool() definitions
|   └── utils.R           # parse_date(), slim_ts(), safe_to_csv()
├── tests
|   ├── testthat.R        # Test framework helper
|   └── testthat
|       ├── test-run.R    # Unit tests of run.R (13 tests)
|       ├── test-tools.R  # Unit tests of tools.R (95 tests)
|       └── test-utils.R  # Unit tests of utils.R (26 tests)
├── man
|   └── run.Rd            # Documentation of functions in R/run.R
└── .github/
    ├── pull_request_template.md
    └── ISSUE_TEMPLATE
       ├── bug_report.md
       └── feature_request.md
```

### Token budget design

Tool results are returned as **CSV** (not JSON) to reduce per-row key repetition, cutting output size 60–70% for wide time-series frames. Constant metadata columns (domain names, unit labels) are dropped automatically by `slim_ts()`. Results are hard-capped at 100 rows with a truncation notice so the LLM always knows when data is partial.

------------------------------------------------------------------------

## Requirements

-   R ≥ 4.2.0
-   [`entsoeapi`](https://github.com/krose/entsoeapi) (GitHub)
-   [`mcptools`](https://github.com/posit-dev/mcptools) (GitHub)
-   `ellmer`, `lubridate`, `jsonlite` (CRAN)

------------------------------------------------------------------------

## License

MIT © Sándor Budai. See [LICENSE](LICENSE).

The underlying data is provided by [ENTSO-E](https://www.entsoe.eu) under their [terms of use](https://transparency.entsoe.eu/content/static_content/Static%20content/terms%20and%20conditions/terms%20and%20conditions.html).
